---
execute: 
  warning: false
  error: false
  message: false
title: "Distance Calculations"
format: 
  html:
    embed-resources: true
    code-fold: true
    code-tools: false
---

# Intro
This document outlines the methods used for geolocating and generating distance calculation between geolocated points. These methods were used in the [The U.S. Department of Housing and Urban Development’s Use and Oversight of Telework and Remote Work](https://www.hudoig.gov/sites/default/files/2024-10/2024-oe-0004_final.pdf) report. 

This example only utilizes data for Flexiplace agreement alternative worksite and and Flexiplace agreement duty station with distances > 500 miles.  

``` {r init}

library(tidygeocoder) 
library(sf)
library(tidyverse)
library(readxl)
library(writexl)
library(geodist)
library(data.table)
library(ggplot2)
library(plotly)

# data 
dat <- read_xlsx("C:/Users/duffinb/Documents/Projects/hud-pay-locality/output/Locality Pay Distance Calculations Request_Dists_XY_2024-04-04.xlsx")
# subset 
dat <- dat |> 
  select(c(`Unique Identifier`, 
           `Distance from Flexiplace Agreement Duty Station to Flexiplace Agreement Alternative Worksite`, 
           FLEXIPLACE_ALTWORKSITE_lat:FLEXIPLACE_DUTY_STATION_lon 
           )) 
names(dat) <- c("ID", "Distance_Mi", "Altsite_lat", "Altsite_lon", 
                "Duty_lat", "Duty_lon")
dat <- subset(dat, 
              Distance_Mi > 500)
dat$grp <- sample(c("a", "b"), nrow(dat), replace = T)
```



# Process 
- Geocode 
  - Inputs: Specific address, or city and state
  - Utilized the [ArcGIS Geocoding Service](https://developers.arcgis.com/rest/geocode/api-reference/overview-world-geocoding-service.htm)
  - Outputs: point locations for each input (for city and state results generally return the centroid)
- Distance Calculation 
  - [Haversine formula](https://en.wikipedia.org/wiki/Haversine_formula) used to calculate distance
  - Inputs: lat/lon endpoints (duty station and alternative worksite in this case)
  - Outputs
  - Convert from meters to miles using conversion factor of 1609

## Other Notes
- There is some error due to the assumption of sphere vs. the real shape of Earth (i.e., a spheroid shape). 
- Haversine w/ about a 0.2% relative error
  - So over 1km there is generally < 1m error
- Great circle distance used as opposed to Rhumb line (constant azimuth)  



# Example 
```{css, echo = FALSE}
.output {
max-height: 800px;
}
```

``` {r greatdist, fig}

geo <- list(
  scope = 'north america',
  projection = list(type = 'azimuthal equal area'),
  showland = TRUE,
  landcolor = toRGB("gray95"),
  countrycolor = toRGB("gray80")
)

geo <- list(
  showland = TRUE,
  showlakes = TRUE,
  showcountries = TRUE,
  showocean = TRUE,
  countrywidth = 0.5,
  landcolor = toRGB("grey90"),
  lakecolor = toRGB("white"),
  oceancolor = toRGB("white"),
  projection = list(
    type = 'orthographic',
    rotation = list(
      lon = -100,
      lat = 40,
      roll = 0
    )
  ),
  lonaxis = list(
   # showgrid = TRUE,
    gridcolor = toRGB("gray40"),
    gridwidth = 0.5
  ),
  lataxis = list(
   # showgrid = TRUE,
    gridcolor = toRGB("gray40"),
    gridwidth = 0.5
  )
)


fig <- plot_geo(locationmode = "USA-states", color = I("red"),
                height = 700)

fig <- fig |> 
  add_markers(data = dat, 
              x = ~Altsite_lon, y = ~Altsite_lat, 
              size = .5) |> 
  add_markers(data = dat, 
              x = ~Duty_lon, y = ~Duty_lat, 
              size = .5)

fig <- fig |> 
  add_segments(
    data = group_by(dat, ID), 
    x = ~Altsite_lon, xend = ~Duty_lon,
    y = ~Altsite_lat, yend = ~Duty_lat,
    alpha = 0.6, size = I(1.3), 
    hoverinfo = "text", 
    text = ~paste0(format(round(Distance_Mi), big.mark = ","), 
                   " Miles")
  ) 

fig <- fig |> 
  layout(title = "Distance Examples", 
         showlegend = FALSE, 
         geo = geo)

fig

```
